SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-09-13 scipy-1.0/paper.tex
Complete outline of SciPy begins section

% Packages
\usepackage[super]{nth}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
\usepackage{pgfplots}

@@

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group
mailing list\footnote{https://mail.python.org/pipermail/matrix-sig/1995-August/000002.html}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
with arbitrarily high dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\footnote{http://hugunin.net/papers/hugunin95numpy.html}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

@@

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.  Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of senior researchers.

% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
Paul Dubois in the Program for Climate Model Diagnosis and
Intercomparison at Lawrence Livermore National Laboratory built their
Climate Data Analysis Tools (``an open-source set of Python modules
and tools for accessing, manipulating, and plotting scientific data in
general and climate data sets in
particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
on Numeric, the newly established Python array object.  Paul later
became the maintainer for Numeric, and helped fund the writing of the
first manual [citation needed].

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 90s, Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\footnote{https://web.archive.org/web/19990125091242/http://oliphant.netpedia.net:80/}\footnote{https://web.archive.org/web/20001206213500/http://oliphant.netpedia.net:80/} that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, and Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\footnote{https://mail.python.org/pipermail/numpy-discussion/2000-September/000243.html}, a command line tool for binding Python and Fortran
codes.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler and genetic optimizer.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical routines\footnote{https://web.archive.org/web/20001022231108/http://www.nmr.mgh.harvard.edu/Neural\_Systems\_Group/gary/python.html}.

In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.

-  Jun 10, 2001
Travis N. Vaught announced scipy-dev mailing list

- Aug 20, 2001
Eric Jones announced \ref{fig:announce-0.1}

IPython 2001 first release

Matplotlib 2003 first release

NumPy 2005

@@

>>>>

% Packages
\usepackage[super]{nth}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
\usepackage{pgfplots}
\usepackage[multiple]{footmisc}

@@

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group (Matrix-SIG)
mailing list\footnote{https://mail.python.org/pipermail/matrix-sig/1995-August/000002.html}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
with arbitrarily high dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\footnote{http://hugunin.net/papers/hugunin95numpy.html}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

@@

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.  Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of senior researchers.
% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
Paul Dubois at Lawrence Livermore National Laboratories took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://people.csail.mit.edu/jrennie/python/numeric/numeric-manual.pdf}.
\fixme{replace this URL with numpy.org hosted version }% Make sure we host this PDF somewhere in the numpy GitHub org
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
In 1998, the Space Telescope Science Institute, in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 90s, discussions started to appear on Matrix-SIG,
expressing a desire for a complete scientific data analysis environment\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\footnote{https://web.archive.org/web/19990125091242/http://oliphant.netpedia.net:80/}\footnote{https://web.archive.org/web/20001206213500/http://oliphant.netpedia.net:80/} that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, and Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\footnote{https://mail.python.org/pipermail/numpy-discussion/2000-September/000243.html}, a command line tool for binding Python and Fortran
codes, and wrote modules for linear algebra and interpolation.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler, genetic optimizer, and inline C++ compiler.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical routines\footnote{https://web.archive.org/web/20001022231108/http://www.nmr.mgh.harvard.edu/Neural\_Systems\_Group/gary/python.html}.

In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.
The new project quickly gained momentum, with a website and code
repository appearing in
February\footnote{https://web.archive.org/web/20010309040805/http://scipy.org:80/}, and
a mailing list announced in
June \footnote{https://mail.python.org/pipermail/scipy-dev/2001-June/000000.html}. By
August, a first release was announced\footnote{https://mail.python.org/pipermail/python-list/2001-August/106419.html}, an excerpt of which is shown in
Fig.~\ref{fig:announce-0.1}.
In September, the first documentation was
published\footnote{https://web.archive.org/web/20021013204556/http://www.scipy.org:80/scipy/site\_content/site\_news/docs\_released1},
and by the following September the inaugural SciPy workshop was held
at Caltech\footnote{https://mail.python.org/pipermail/numpy-discussion/2002-June/001511.html}.

% Janko Hauser on toolboxes
% https://mail.python.org/pipermail/scipy-dev/2001-September/000032.html

At this point, scientific Python is starting to attract more serious
attention; code that started out as side projects by graduate students
has grown into essential infrastructure at national laboratories and
research institutes.  But even as SciPy, the algorithms library of the
ecosystem, started to mature, the next generation of graduate students
and postdocs were already exploring other aspects of the computational
environment.

In 2000, Prabhu Ramachandran\fixme{also note that he, and the others
  mentioned below, are also SciPy contributors}, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.

SciPy initially contained a plotting module, but it did not have an
easy and familiar interface. \fixme{There were multiple plotting
  packages in SciPy, and Chaco, and wxPyPlot(s), IPython + GNUPlot,
  and ....?}
John Hunter, a gradute student
\fixme{check, since John around that time became postdoc, can check
  with John Milton?} at the University
of Chicago Medical School, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell, built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html

In 2000, STScI ported their entire IDL-based analysis pipeline, IRAF,
to Python.  And it was here that a small crisis developed around the
Python numerical array container.  Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI.  With the Numeric maintainer's blessing, the decision was made
to write NumArray, a library that could handle data on a different
scale.  Unfortunately, NumArray proved inefficient for small arrays,
facing the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement. \fixme{mention
  first NumPy meeting at Berkeley}

% Future directions for SciPy after meeting at Berkeley
% https://mail.python.org/pipermail/scipy-user/2005-March/004164.html

@@

% While Paul was one of the early adopters at the national laboratories,
% Python's presence is now ubiquitous.
